---
title: "R4DS Notes and Exercises"
author: "Eric Allatta"
date: "03/22/2026"
---

Exercises and examples from [R 4 Data Science](https://r4ds.hadley.nz/) and [R for Everyone](https://www.jaredlander.com/r-for-everyone/), combined and grouped by topic and package.

## setup

```{r}
library('here')
```

# data.table

https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html
Everyone says this is super fast package, but you need to understand some other style than tidy.
Let's take a look.

```{r}
library('data.table')
```

```{r}
download.file('https://raw.githubusercontent.com/Rdatatable/data.table/master/vignettes/flights14.csv',destfile=here('data/flights14.csv'))
```

```{r}
input <- here('data/flights14.csv')
flights <- fread(input)
flights
```

fread is pretty cool. You can give it just about anything.

`?fread`

## basics

data.table is (provides) an enhanced data.frame.
fread or call data.table() function

```{r}
DT = data.table(
    ID = c('b', 'b', 'b', 'a', 'a', 'c'),
    a = 1.6,
    b = 7:12,
    c = 13:18
)
DT
```

```{r}
class(DT$ID)
```

See ?setDT and ?as.data.table for converting existing objects to data.table
data.table doesn't set or use *row names* ever. Something about this is why it's so fast.
[See](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-keys-fast-subset.html)

### enhanced?
a lot can happen inside DT[ ... ], this more like querying DT analog to SQL

### general form
DT[i, j, by]

### subset rows
Get all the flights with 'JFK' as the origin airport in the month of June.

```{r}
##   R:                i                 j        by
## SQL: where | order by   select | update  group by
```

Take DT, subset/reorder rows using i, then calculate j, group by `by`.

Refer to the columns as if they are variables!
Check out `origin` and `month`

```{r}
ans <- flights[origin == 'JFK' & month == 6L]
head(ans)
```

*get the first two rows from flights*

```{r}
ans <- flights[1:2]
ans
```

*Sort `flights` by  column*
Sounds like order is pretty cool, base::R switch to using its algorithm
```{r}
# sort by origin and reverse sort by dest
ans <- flights[order(origin, -dest)]
head(ans)
```
### select and comput columns

```{r}
# select arr_delay column, but return as a data.table instead
ans <- flights[, list(arr_delay)] # wrapping in list ensure data.table is returned
head(ans)
```
`.()` is an *alias* to `list()`
data.table and data.frame are internally `list`. every element has to be the same lenght and the `list` has a `class` attribute

```{r}
# select both arr_delay and dep_delay columns
ans <- flights[, .(arr_delay, dep_delay)]
head(ans)
```

**select and rename**

```{r}
ans <- flights[, .(delay_arr = arr_delay, delay_dep = dep_delay)]
head(ans)
```

### compute or do in j

Sum the number of trips that have delays < 0. Functions expressed directly on columns as variables.

```{r}
ans <- flights[, sum( (arr_delay + dep_delay) < 0 )]
ans
```

### subset in `i` *and* do in `j`

```{r}
ans <- flights[origin == 'JFK' & month == 6L,
               .(m_arr = mean(arr_delay), m_dep = mean(dep_delay))]
ans
```

### perform data aggregatios by group

# Chapter 5 - Data tidying
```{r}
#| echo: true
#| eval: true
#| output: false
library('tidyverse')
```

```{r}
billboard
```
Billboard has 79 columns, mostly wk1, wk2. Needs to be flattened, i.e. pivot_longer()
Similar to program groups in hs_directory, but only one group: the week.

In tidy data each column is a variable. Week is the variable, so one column per week is not a variable.

```{r}
billboard %>%
    pivot_longer(
        cols = starts_with('wk'), # specifies which columns need to be pivoted
        names_to = 'week', # names the variable stored in the column name
        values_to = 'rank', # the value stored in the cell values, new variables quoted because they don't exist yet
        values_drop_na = TRUE # because missing data was forced by the structure, not having data for a week means you weren't on the chart, not that data is missing
    ) %>%
        mutate(
            week = parse_number(week) # this is a new column; parse_number is a very powerful parser that extracts just the number
        ) -> billboard_longer
billboard_longer
```

```{r}
billboard_longer %>%
    ggplot(aes(x = week, y = rank, group = track)) +
    geom_line(alpha = 0.25) +
    scale_y_reverse()
```
